% =============================================================================
%  Imperial College London — MEng (AI & Machine Learning) Final Year Project
%  Structured Refactoring Agent for IntelliJ
%
%  Build:  cd report && latexmk -pdf main.tex     (or pdflatex x2 + bibtex)
%  This is a SCAFFOLD. Prose marked [WRITE: ...] is yours to author.
%  See WRITING_GUIDE.md for the page budget, rubric mapping, and day-by-day.
% =============================================================================
\documentclass[11pt,a4paper]{report}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}   % scalable fonts (required by microtype expansion)
\usepackage[a4paper,margin=2.5cm]{geometry}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{microtype}
\usepackage[numbers,square]{natbib}
\usepackage{hyperref}
\usepackage{cleveref}   % load AFTER hyperref

\graphicspath{{figures/}}

% ---- Visible placeholder for content you still need to write -----------------
\newcommand{\writeme}[1]{\textcolor{red}{\textbf{[WRITE:} #1\textbf{]}}}

% ---- Code listing style ------------------------------------------------------
\definecolor{codegray}{gray}{0.95}
\lstdefinestyle{code}{
  backgroundcolor=\color{codegray},
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  captionpos=b,
  frame=single,
  numbers=left,
  numberstyle=\tiny,
  showstringspaces=false,
  tabsize=2,
}
\lstset{style=code}

% ---- Project metadata (FILL THESE IN) ---------------------------------------
\newcommand{\reporttitle}{Structured Refactoring Agent for IntelliJ:\\Decoupling LLM Reasoning from AST-Safe Execution}
\newcommand{\reportauthor}{Ibrahim Ibrahim}
\newcommand{\supervisorname}{Dr Robert Chatley}
\newcommand{\secondmarkername}{\writeme{second marker name}}
\newcommand{\reportdate}{June 2026}

% NOTE: PDF metadata must be plain text — no \\ and no \writeme/\textcolor here.
\hypersetup{
  colorlinks=true, linkcolor=black, citecolor=black, urlcolor=blue,
  pdftitle={Structured Refactoring Agent for IntelliJ},
  pdfauthor={Ibrahim}
}

\begin{document}

% ============================== TITLE PAGE ===================================
\begin{titlepage}
  \centering
  \vspace*{1cm}
  {\scshape\LARGE Imperial College London \par}
  \vspace{0.3cm}
  {\scshape\large Department of Computing \par}
  \vspace{2.5cm}
  {\huge\bfseries \reporttitle \par}
  \vspace{2.5cm}
  {\Large \reportauthor \par}
  \vspace{0.2cm}
  {\large CID: 01870877 \par}
  \vspace{1cm}
  {\large Supervisor: \supervisorname \par}
  {\large Second Marker: \secondmarkername \par}
  \vfill
  {\large Submitted in partial fulfilment of the requirements for the\\
   MEng degree in Computing (Artificial Intelligence and Machine Learning)\par}
  \vspace{1cm}
  {\large \reportdate \par}
\end{titlepage}

% ============================== ABSTRACT =====================================
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
% ~250 words. Write this LAST, once results are final.
% Must state: problem, what you built, what you measured, the headline result.
Large language models can reliably identify useful refactorings but apply them
unreliably: when an LLM rewrites source text directly it misses cross-file usages,
breaks imports, and produces changes that compile yet are semantically wrong, with the
strongest models reaching only about 41.6\% on repository-level refactoring benchmarks.
This project argues that the limitation lies in the \emph{action space} rather than the
model, and addresses it by decoupling reasoning from execution. An IntelliJ~IDEA plugin
exposes the IDE's PSI-backed refactoring engine --- rename, move, change-signature,
safe-delete, inline, extract, and symbol creation --- as a model-agnostic,
externally-callable tool API, over HTTP and the Model Context Protocol, that any
function-calling agent can drive. Because every edit is performed by the IDE engine
rather than emitted as text, each change is behaviour-preserving and propagated across
the whole codebase by construction.

The system is evaluated with a controlled comparison that holds the model fixed and
varies only the edit primitive, across three real-world Java projects (spring-petclinic,
Apache Commons Lang, and jackson-databind, up to 474 source files), scored by
compilation and disk-state checks. The result is deliberately nuanced: against scripted
and weak-model text baselines the structured engine wins clearly (9/9 versus 6/9 on
spring-petclinic), but a \emph{capable} agent editing raw text reaches parity on
behaviour-preserving refactorings (12 of 12 across the three projects). The structured
action space's durable, measured advantages are therefore correctness \emph{by
construction} --- independent of the model's capability and of any build-feedback loop
--- exhaustive, type-aware structural queries that text editing cannot perform at all,
and far fewer mutating operations; advantages that prove decisive precisely where the
driving agent is weak, scripted, or unverified.

% ============================== DECLARATION ==================================
\chapter*{Declarations}
\addcontentsline{toc}{chapter}{Declarations}

\noindent\textbf{Originality.}\\
% This must be YOUR honest statement. Template only.
This report and the work it describes are my own, except where explicitly stated and
referenced.

\vspace{1em}
\noindent\textbf{Source code.}\\
The implementation is available at:
\url{https://github.com/ibrahim-io/structured-refactoring-agent-intellij}.
% NOTE (rubric requirement): the repo link goes here AND your supervisor must
% have been granted access to the repository.

\vspace{1em}
\noindent\textbf{Use of generative AI.}\\
% CRITICAL — see WRITING_GUIDE.md. Undeclared/over-relied-upon AI => 0-39% cap
% + possible misconduct referral. Acceptable use (coding assistance, structuring,
% proofreading) is fine IF disclosed, human-reviewed, and you remain the author.
% Write this honestly and specifically. Template prompts below — REPLACE them.
This project made substantial and disclosed use of generative AI --- specifically
Anthropic's Claude (the Claude Opus~4.8 model, used through the Claude Code agentic
command-line tool) --- in three roles. First, as a \emph{coding assistant} for parts of
the plugin and the evaluation harness, including the headless-execution technique and
the benchmark runners. Second, in \emph{writing}: to produce first drafts and structure
for this report from my own research notes, design decisions, and measured results,
which I then reviewed, corrected, rewrote where necessary, and verified against the
source code and the data. Third, in the \emph{evaluation itself}: Claude was used both
to orchestrate the benchmark runs and, in the controlled comparison
(\Cref{sec:eval-controlled}), as the ``capable agent'' under test that drives the
refactoring tools; this dual role is stated explicitly in the methodology, and means
that arm demonstrates the system's intended deployment rather than an independent
third-party benchmark. I directed the work throughout, I checked the factual claims,
figures, and citations myself, and I am the author of both the system and this report;
all AI-generated text and code was reviewed and adapted by me rather than submitted as
produced, and I can explain and defend every part.

\vspace{1em}
\noindent\textbf{Ethics.}\\
This project involved neither human participants nor personal data: it is a software
system evaluated entirely on open-source code repositories. The developer study
originally considered as an extension was not conducted, so the project required no
ethics approval.

% ============================ ACKNOWLEDGEMENTS ================================
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
I am grateful to my supervisor, Dr Robert Chatley, for his guidance and feedback
throughout this project, and for the freedom to pursue the direction it took. I
would also like to thank my family and friends for their support over the course
of my degree at Imperial College London.

% ============================== CONTENTS =====================================
\tableofcontents
\listoffigures
\listoftables

% ============================== CHAPTERS =====================================
\input{chapters/01-introduction}
\input{chapters/02-background}
\input{chapters/03-design}
\input{chapters/04-implementation}
\input{chapters/05-evaluation}
\input{chapters/06-conclusion}

% ============================== BIBLIOGRAPHY =================================
\bibliographystyle{plainnat}
\bibliography{references}

% ============================== APPENDICES ===================================
\appendix
\chapter{Tool API Reference}
% Full 24-tool JSON schema, example requests/responses. Good place to move
% detail out of the Implementation chapter to respect the page limit
% (check whether appendices count toward your page limit — see WRITING_GUIDE.md).
The tool surface is published as a single JSON schema served at
\texttt{GET /tools/schema} (the live schema is authoritative; a grouped summary appears
in \Cref{tab:tools}). Each tool is invoked by POSTing a \texttt{\{"tool",\,"params"\}}
body to \texttt{/tools}. Two representative calls follow.

A rename by qualified name:
\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
POST /tools
{"tool": "rename_symbol",
 "params": {"qualifiedName": "...petclinic.owner.Owner#telephone",
            "newName": "phoneNumber"}}
=> {"ok": true, "message": "renamed telephone to phoneNumber"}
\end{lstlisting}

A project-wide, type-aware usage query (no text-level equivalent):
\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
POST /tools
{"tool": "find_usages",
 "params": {"qualifiedName": "org.apache.commons.lang3.CharRange"}}
=> {"count": 213, "usages": [{"filePath": "...", "line": 61, "preview": "..."}]}
\end{lstlisting}
The complete operation set, parameters, and return shapes are documented in the live
schema and in the project repository.

\chapter{Benchmark Task Definitions}
Each task in the benchmark suites (\texttt{benchmarks/tasks\_petclinic.json},
\texttt{tasks\_commons\_lang.json}, and \texttt{tasks\_jackson.json}) carries a
natural-language description, the canonical operation(s), and a machine-checkable success
specification. One representative entry --- the overload-safe rename of
\Cref{sec:eval-rootcause} --- is:
\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
{"id": "pc-rename-method-002",
 "description": "Rename the two-argument overload getPet(String, boolean) in Owner
                 to findPet; do not rename the other getPet overloads.",
 "project": "spring-petclinic",
 "operations": [{"tool": "rename_symbol",
                 "params": {"qualifiedName": "...owner.Owner#getPet(String,boolean)",
                            "newName": "findPet"}}],
 "validation": {"type": "compile_and_symbol_exists",
                "expectedSymbol": "findPet", "crossFileCheck": true}}
\end{lstlisting}
The full suites are available in the project repository.

\end{document}
